# Manipulating tree with data


```{r include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
```



```{r echo=FALSE, results="hide", message=FALSE}
library("ape")
library("tidytree")
```


Manipulating tree object is frustrated with the fragmented functions available
for working with `phylo` object, not to mention linking external data to the
phylogeny structure.
Using [tidy data principles](https://www.jstatsoft.org/article/view/v059i10) can
make phylogenetic tree manipulation tasks easier and consistent with tools
already in wide use,
including
[dplyr](https://cran.r-project.org/package=dplyr),
[tidyr](https://cran.r-project.org/package=tidyr),
[ggplot2](https://cran.r-project.org/package=ggplot2)
and [ggtree](https://bioconductor.org/packages/ggtree/).


## Convert tree object to tidy data frame and vice versa

### `phylo` object


The `phylo` class defined in [ape](https://cran.r-project.org/package=ape) is
fundamental for phylogenetic analysis in R. Most of the R packages
in [this field](https://CRAN.R-project.org/view=Phylogenetics) rely
extensively on `phylo` object. The tidytree package provides `as_tibble`
method to convert the `phylo` object to tidy data frame, a `tbl_tree` object.

```{r}
library(ape)
set.seed(2017)
tree <- rtree(4)
tree
x <- as_tibble(tree)
x
```

The `tbl_tree` object can be converted back to a `phylo` object.


```{r}
as.phylo(x)
```


Using `tbl_tree` object makes tree and data manipulation more effective and
easier. For example, we can link evolutionary trait to phylogeny using the verbs `full_join`

```{r}
d <- tibble(label = paste0('t', 1:4),
            trait = rnorm(4))

y <- full_join(x, d, by = 'label')
```

### treedata object


The tidytree package defines a `treedata` class to store phylogenetic tree with
associated data. After mapping external data to the tree structure, the
`tbl_tree` object can be converted to a `treedata` object.


```{r}
as.treedata(y)
```

The `treedata` class is also used
in [treeio](https://bioconductor.org/packages/treeio/) package to store
evolutionary evidences inferred by commonly used software (BEAST, EPA, HYPHY,
MrBayes, PAML, PHYLODOG, pplacer, r8s, RAxML and RevBayes).

The tidytree package also provides `as_tibble` to convert `treedata` object
to a tidy data frame. The phylogentic tree structure and the evolutionary
inferences were stored in the tbl_tree object, making it consistent and easier
for manipulating evolutionary statistics inferred by different software as well
as linking external data to the same tree structure.


```{r}
y %>% as.treedata %>% as_tibble
```


## Access related nodes

[dplyr](https://cran.r-project.org/package=dplyr) verbs can be applied to `tbl_tree` directly to manipulate tree data. In
addition, tidytree provides several verbs to filter related nodes, including
`child`, `parent`, `offspring`, `ancestor`, `sibling` and `MRCA`.

These verbs accept a `tbl_tree` and a selected node which can be node number or label.


```{r}
child(y, 5)
parent(y, 2)
offspring(y, 5)
ancestor(y, 2)
sibling(y, 2)
MRCA(y, 2, 3)
```

## Grouping taxa

tidytree implemented `groupOTU` and `groupClade` for adding taxa grouping
information to the input `tbl_tree` object. These grouping information can be
used directly in tree visualization (*e.g.* coloring tree based on grouping)
with [ggtree](https://bioconductor.org/packages/ggtree/).

### groupClade

The `groupClade` method accepts an internal node or a vector of internal nodes
to add grouping information of clade/clades.



```{r}
nwk <- '(((((((A:4,B:4):6,C:5):8,D:6):3,E:21):10,((F:4,G:12):14,H:8):13):13,((I:5,J:2):30,(K:11,L:11):2):17):4,M:56);'
tree <- read.tree(text=nwk)

groupClade(as_tibble(tree), c(17, 21))
```


### groupOTU


```{r}
## the input nodes can be node ID or label
groupOTU(x, c('t1', 't4'), group_name = "fake_group")
```

The `groupOTU` will trace back from input nodes to most recent common ancestor.
In this example, nodes 2, 3, 7 and 6 (`2 (t1) -> 7 -> 6` and `3 (t4) -> 6`) are
grouping together.

Related OTUs are grouping together and they are not necessarily within a clade.
They can be monophyletic (clade), polyphyletic or paraphyletic.

```{r}
cls <- list(c1=c("A", "B", "C", "D", "E"),
            c2=c("F", "G", "H"),
            c3=c("L", "K", "I", "J"),
            c4="M")

as_tibble(tree) %>% groupOTU(cls)
```

If there are conflicts when tracing back to mrca, user can set `overlap`
parameter to "origin" (the first one counts), "overwrite" (default, the last one
counts) or "abandon" (un-selected for grouping), see
also
[discussion here](https://groups.google.com/forum/#!msg/bioc-ggtree/Q4LnwoTf1DM/uqYdYB_VBAAJ).


