# Phylogenetic Tree Annotation {#chapter5}


```{r include=FALSE}
library("ape")
library("ggplot2")
library("cowplot")
library("treeio")
library("ggtree")


inset <- ggtree::inset
```


## Visualizing and Annotating Tree using Grammar of Graphics

The *ggtree* is designed for more general purpose or specific type of tree visualization and annotation. It supports grammar of graphics\index{grammar of graphics} implemented in *ggplot2* and users can freely visualize/annotate a tree by combining several annotation layers. 

(ref:ggtreeNHXscap) Annotating tree using grammar of graphics.

(ref:ggtreeNHXcap) **Annotating tree using grammar of graphics.** The NHX tree was annotated using grammar of graphic syntax by combining different layers using `+` operator. Species information were labelled on the middle of the branches, Duplication events were shown on most recent common ancestor and clade bootstrap value were dispalyed near to it.


```{r echo=F, message=F, warning=F}
library(ggtree)
treetext = "(((ADH2:0.1[&&NHX:S=human], ADH1:0.11[&&NHX:S=human]):
0.05 [&&NHX:S=primates:D=Y:B=100],ADHY:
0.1[&&NHX:S=nematode],ADHX:0.12 [&&NHX:S=insect]):
0.1[&&NHX:S=metazoa:D=N],(ADH4:0.09[&&NHX:S=yeast],
ADH3:0.13[&&NHX:S=yeast], ADH2:0.12[&&NHX:S=yeast],
ADH1:0.11[&&NHX:S=yeast]):0.1[&&NHX:S=Fungi])[&&NHX:D=N];"
tree <- read.nhx(textConnection(treetext))
p = ggtree(tree) + geom_tiplab() + 
  geom_label(aes(x=branch, label=S), fill='lightgreen') + 
  geom_label(aes(label=D), fill='steelblue') + 
  geom_text(aes(label=B), hjust=-.5)
p <- p + xlim(NA, 0.28)
```

\setstretch{1.2}

```{r echo=T, eval=F}
library(ggtree)
treetext = "(((ADH2:0.1[&&NHX:S=human], ADH1:0.11[&&NHX:S=human]):
0.05 [&&NHX:S=primates:D=Y:B=100],ADHY:
0.1[&&NHX:S=nematode],ADHX:0.12 [&&NHX:S=insect]):
0.1[&&NHX:S=metazoa:D=N],(ADH4:0.09[&&NHX:S=yeast],
ADH3:0.13[&&NHX:S=yeast], ADH2:0.12[&&NHX:S=yeast],
ADH1:0.11[&&NHX:S=yeast]):0.1[&&NHX:S=Fungi])[&&NHX:D=N];"
tree <- read.nhx(textConnection(treetext))
ggtree(tree) + geom_tiplab() + 
  geom_label(aes(x=branch, label=S), fill='lightgreen') + 
  geom_label(aes(label=D), fill='steelblue') + 
  geom_text(aes(label=B), hjust=-.5)
```

```{r ggtreeNHX, warning=FALSE, fig.cap="(ref:ggtreeNHXcap)", fig.scap="(ref:ggtreeNHXscap)", out.extra='', echo=F}
print(p)
```
\setstretch{1.5}

Here, as an example, we visualized the tree with several layers to display annotation stored in NHX tags, including a layer of `geom_tiplab` to display tip labels (gene name in this case), a layer using `geom_label` to show species information (`S` tag) colored by lightgreen, a layer of duplication event information (`D` tag) colored by steelblue and another layer using *geom_text* to show bootstrap value (`B` tag).

Layers defined in *ggplot2* can be applied to *ggtree* directly as demonstrated in Figure \@ref(fig:ggtreeNHX) of using *geom_label* and *geom_text*. But *ggplot2* does not provide graphic layers that are specific designed for phylogenetic tree annotation. For instance, layers for tip labels, tree branch scale legend, highlight or labeling clade are all unavailable. To make tree annotation more flexible, a number of layers have been implemented in *ggtree* (Table \@ref(tab:geoms)), enabling different ways of annotation on various parts/components of a phylogenetic tree.

```{r geoms, echo=FALSE, message=FALSE}
geoms <- matrix(c(
  "geom_balance", "highlights the two direct descendant clades of an internal node",
  "geom_cladelabel", "annotate a clade with bar and text label",
  "geom_hilight", "highlight a clade with rectangle",
  "geom_label2", "modified version of geom_label, with subsetting supported",
  "geom_nodepoint", "annotate internal nodes with symbolic points",
  "geom_point2", "modified version of geom_point, with subsetting supported",
  "geom_range", "bar layer to present uncertainty of evolutionary inference",
  "geom_rootpoint", "annotate root node with symbolic point",
  "geom_segment2", "modified version of geom_segment, with subsetting supported",
  "geom_strip", "annotate associated taxa with bar and (optional) text label",
  "geom_taxalink", "associate two related taxa by linking them with a curve",
  "geom_text2", "modified version of geom_text, with subsetting supported",
  "geom_tiplab", "layer of tip labels",
  "geom_tiplab2", "layer of tip labels for circular layout",
  "geom_tippoint", "annotate external nodes with symbolic points",
  "geom_tree", "tree structure layer, with multiple layout supported",
  "geom_treescale", "tree branch scale legend"
), ncol=2, byrow=TRUE)
geoms <- as.data.frame(geoms)
colnames(geoms) <- c("Layer", "Description")
knitr::kable(geoms, caption = "Geom layers defined in ggtree.", booktabs = T) %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = T)
```

## Annotate clades

`r Biocpkg("ggtree")` [@yu_ggtree:_2017] implements _`geom_cladelabel`_ layer to annotate a selected clade with a bar indicating the clade with a corresponding label.

The _`geom_cladelabel`_ layer accepts a selected internal node number. To get the internal node number, please refer to [Tree Manipulation](treeManipulation.html#internal-node-number) vignette.


```{r}
set.seed(2015-12-21)
tree <- rtree(30)
p <- ggtree(tree) + xlim(NA, 6)

p + geom_cladelabel(node=45, label="test label") +
    geom_cladelabel(node=34, label="another clade")
```

Users can set the parameter, `align = TRUE`, to align the clade label, and use the parameter, `offset`, to adjust the position.

```{r}
p + geom_cladelabel(node=45, label="test label", align=TRUE, offset=.5) +
    geom_cladelabel(node=34, label="another clade", align=TRUE, offset=.5)
```

Users can change the color of the clade label via the parameter `color`.

```{r}
p + geom_cladelabel(node=45, label="test label", align=T, color='red') +
    geom_cladelabel(node=34, label="another clade", align=T, color='blue')
```

Users can change the `angle` of the clade label text and relative position from text to bar via the parameter `offset.text`.

```{r}
p + geom_cladelabel(node=45, label="test label", align=T, angle=270, hjust='center', offset.text=.5) +
    geom_cladelabel(node=34, label="another clade", align=T, angle=45)
```

The size of the bar and text can be changed via the parameters `barsize` and `fontsize` respectively.

```{r}
p + geom_cladelabel(node=45, label="test label", align=T, angle=270, hjust='center', offset.text=.5, barsize=1.5) +
    geom_cladelabel(node=34, label="another clade", align=T, angle=45, fontsize=8)
```

Users can also use `geom_label` to label the text.

```{r}
p + geom_cladelabel(node=34, label="another clade", align=T, geom='label', fill='lightblue')
```

### Annotate clades for unrooted tree

`r Biocpkg("ggtree")` provides `geom_clade2` for labeling clades of unrooted
layout trees.


```{r fig.wdith=7, fig.height=7, fig.align='center', warning=FALSE, message=FALSE}
pg <- ggtree(tree, layout="daylight")
pg + geom_cladelabel2(node=45, label="test label", angle=10) +
    geom_cladelabel2(node=34, label="another clade", angle=305)
```

## Labelling associated taxa (Monophyletic, Polyphyletic or Paraphyletic)

`geom_cladelabel` is designed for labelling Monophyletic (Clade) while there are related taxa that are not form a clade. `ggtree` provides `geom_strip` to add a strip/bar to indicate the association with optional label (see [the issue](https://github.com/GuangchuangYu/ggtree/issues/52)).

```{r fig.width=5, fig.height=5, fig.align="center", warning=FALSE}
nwk <- system.file("extdata", "sample.nwk", package="treeio")
tree <- read.tree(nwk)
ggtree(tree) + geom_tiplab() + 
  geom_strip(5, 7, barsize=2, color='red') + 
  geom_strip(6, 12, barsize=2, color='blue')
```


## Highlight clades

`ggtree` implements _`geom_hilight`_ layer, that accepts an internal node number and add a layer of rectangle to highlight the selected clade.

```{r fig.width=5, fig.height=5, fig.align="center", warning=FALSE}
ggtree(tree) + geom_hilight(node=21, fill="steelblue", alpha=.6) +
    geom_hilight(node=17, fill="darkgreen", alpha=.6)
```


```{r fig.width=5, fig.height=5, fig.align="center", warning=FALSE}
ggtree(tree, layout="circular") + geom_hilight(node=21, fill="steelblue", alpha=.6) +
    geom_hilight(node=23, fill="darkgreen", alpha=.6)
```

Another way to highlight selected clades is setting the clades with different colors and/or line types as demonstrated in [Tree Manipulation](treeManipulation.html#groupclade) vignette.

### Highlight balances

In addition to _`geom_hilight`_, `ggtree` also implements _`geom_balance`_
which is designed to highlight neighboring subclades of a given internal node.

```{r fig.width=4, fig.height=5, fig.align='center', warning=FALSE}
ggtree(tree) +
  geom_balance(node=16, fill='steelblue', color='white', alpha=0.6, extend=1) +
  geom_balance(node=19, fill='darkgreen', color='white', alpha=0.6, extend=1)
```

### Highlight clades for unrooted tree

`r Biocpkg("ggtree")` provides `geom_hilight_encircle` to support highlight
clades for unrooted layout trees.


```{r fig.width=5, fig.height=5, fig.align='center', warning=FALSE, message=FALSE}
pg + geom_hilight_encircle(node=45) + geom_hilight_encircle(node=34, fill='darkgreen')
```


## Taxa connection

Some evolutionary events (e.g. reassortment, horizontal gene transfer) can be modeled by a simple tree. `ggtree` provides `geom_taxalink` layer that allows drawing straight or curved lines between any of two nodes in the tree, allow it to represent evolutionary events by connecting taxa.

```{r fig.width=5, fig.height=5, fig.align="center", warning=FALSE}
ggtree(tree) + geom_tiplab() + geom_taxalink('A', 'E') + 
  geom_taxalink('F', 'K', color='red', arrow=grid::arrow(length=grid::unit(0.02, "npc")))
```


## Tree annotation with output from evolution software

The `r Biocpkg("treeio")` package implemented several parser functions to parse
output from commonly used software in evolutionary biology.

Here, we used [BEAST](http://beast2.org/) [@bouckaert_beast_2014] output as an
example. For details, please refer to the
[Importer](https://bioconductor.org/packages/devel/bioc/vignettes/treeio/inst/doc/Importer.html) vignette.


```{r warning=FALSE, fig.width=5, fig.height=5, fig.align='center'}
file <- system.file("extdata/BEAST", "beast_mcc.tree", package="treeio")
beast <- read.beast(file)
ggtree(beast, aes(color=rate))  +
    geom_range(range='length_0.95_HPD', color='red', alpha=.6, size=2) +
    geom_nodelab(aes(x=branch, label=round(posterior, 2)), vjust=-.5, size=3) +
    scale_color_continuous(low="darkgreen", high="red") +
    theme(legend.position=c(.1, .8))
```


