# (PART\*) Part III: Miscellaneous topics {-}

# ggtree utilities {#chapter9}

## facet_widths

```{r}
library(ggplot2)
library(ggstance)
library(ggtree)
library(reshape2)

set.seed(123)
tree <- rtree(30)

p <- ggtree(tree, branch.length = "none") + 
    geom_tiplab()

a <- runif(30, 0,1)
b <- 1 - a
df <- data.frame(tree$tip.label, a, b)
df <- melt(df, id = "tree.tip.label")

p2 <- facet_plot(p + xlim_tree(8), panel = 'bar', data = df, geom = geom_barh, 
                 mapping = aes(x = value, fill = as.factor(variable)), 
                 width = 0.8, stat='identity')

facet_widths(p2, widths = c(1, 2))
```

It also supports using name vector to set the widths of specific panels.

```{r}
facet_widths(p2, c(Tree = .5))
```


The `facet_widths` function also work with other `ggplot` object.

```{r include=FALSE}
theme_set(theme_grey())
```

```{r}
p <- ggplot(iris, aes(Sepal.Width, Petal.Length)) + 
  geom_point() + facet_grid(.~Species)
facet_widths(p, c(setosa = .5))
```

## facet_labeller

The `facet_labeller` function was designed to re-label selected panels, and it currently only works with `ggtree` object (*i.e.* `facet_plot` output).

```{r}
facet_labeller(p2, c(Tree = "phylogeny", bar = "HELLO"))
```

If you want to combine `facet_widths` with `facet_labeller`, you need to call `facet_labeller` to re-label the panels before using `facet_widths` to set the relative widths of each panels. Otherwise it wont work since the output of `facet_widths` is re-drawn from `grid` object.

```{r}
facet_labeller(p2, c(Tree = "phylogeny")) %>% facet_widths(c(Tree = .4))
```

