
# Other ggtree extensions {#chapter11}

```{r include=F}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
```

```{r revdep}
rp <- BiocManager::repositories()
db <- utils::available.packages(repo=rp)
x <- tools::package_dependencies('ggtree', db=db, 
                                which = c("Depends", "Imports"), 
                                reverse=TRUE)
print(x)                                
```

There are `r length(unlist(x))` packages in CRAN or Bioconductor that depend or import `r Biocpkg("ggtree")` and several packages on github that extends ggtree.



+ [deeptime](https://github.com/willgearty/deeptime): Plotting Tools for Anyone Working in Deep Time
+ [ggnetworx](https://github.com/KlausVigo/ggnetworx): phylogenetic networks using ggplot2 and ggtree
+ [ggdaisie](https://github.com/richelbilderbeek/ggdaisie): ggtree and DAISIE R package
+ [MicrobiotaProcess](https://github.com/YuLab-SMU/MicrobiotaProcess): an package for analysis, visualization and biomarker discovery of microbiome 
+ [microbiomeViz](https://github.com/lch14forever/microbiomeViz): Visualize microbiome data with black magic ggtree


