# (APPENDIX) Appendix {-}

# Image processing using magick package

`r CRANpkg("ggimage")` allows processing images on the fly via the `image_fun` parameter, which accepts a function to process `magick-image` object. The `r CRANpkg("magick")` package provides several functions and these functions can be combined to perform a particular task.

## Example 1: Remove background of images

```{r ggimage-background, fig.width=8, fig.height=4} 
set.seed(1982)
x <- rtree(5)
p <- ggtree(x) + theme_grey()
p1 <- p + geom_nodelab(image="img/frogs/frog.jpg", geom="image", size=.12) + 
          ggtitle("original image")
p2 <- p + geom_nodelab(image="img/frogs/frog.jpg", geom="image", size=.12,
            image_fun= function(.) magick::image_transparent(., "white")) + 
            ggtitle("image with background removed")
plot_grid(p1, p2, ncol=2)
```


# Figures and Tables

```{r treeio, echo=F, message=FALSE}
library(kableExtra)
ff <- matrix(c(
  'read.beast',  "parsing output of BEAST",
  "read.codeml", "parsing output of CodeML (rst and mlc files)",
  "read.codeml_mlc", "parsing mlc file (output of CodeML)",
  "read.hyphy", "parsing output of HYPHY",
  "read.jplace", "parsing jplace file including output of EPA and pplacer",
  "read.nhx", "parsing NHX file including output of PHYLODOG and RevBayes",
  "read.paml_rst", "parsing rst file (output of BaseML or CodeML)",
  "read.phylip", "parsing phylip file",
  "read.phyloT", "parsing output of phyloT (http://phylot.biobyte.de/)",
  "read.r8s", "parsing output of r8s",
  "read.raxml", "parsing output of RAxML",
  "Nnode", "get number of internal nodes",
  "Nnode2", "get number of all nodes (including internal and external nodes)",
  "Ntip", "get number of all tips (external nodes)",
  "as.phylo", "convert tree object to phylo object",
  "as.treedata", "convert tree object to treedata object",
  "drop.tip", "remove tips in a tree",
  "get.fields", "get available tree attributes stored in tree object",
  "get.placements", "get placement result of jplace object",
  "get.subs", "get substitution by comparing parent node and child node",
  "get.tipseq", "get tip sequences",
  "get.treetext", "get newick text of a tree object",
  "groupClade", "add clade grouping information to tree object",
  "groupOTU", "grouping OTUs by tracing back to most recent commond ancestor",
  "merge_tree", "merging tree objects into one",
  "write.jpace", "write tree with associated annotation data to a jplace file"
  ), ncol=2, byrow=TRUE)
ff <- as.data.frame(ff)
colnames(ff) <- c("Function", "Description")
knitr::kable(ff, caption = "Functions defined in treeio.", booktabs = T) #%>%
  #kable_styling(latex_options = c("striped", "hold_position"), full_width = T)
```



