
## Software information and conventions {-}

```{r include=FALSE}
#options(width = 30)

source('conflicted.R')
source("software-link.R")
```

The R and core packages information when compiling this book is as follows:

```{r tidy=FALSE, message=TRUE}
R.version.string

library(treedataverse)
```

The `r pkg_treedataverse` is a meta package to make it easy to install and load core packages for processing and visualizing tree with data using the packages described in this book. The installation guide for `r pkg_treedataverse` can be found in FAQ. 

The datasets used in this book have three sources:

1. Simulation data
2. Datasets in the R packages
3. Data downloaded from the Internet

In order to make the data downloaded from the Internet more accessible, we packed the data in an R package, `r CRANpkg("TDbook")`, with detailed documentation of the original source, including URL, authors, and citation if the information is available. The `r CRANpkg("TDbook")` is available on CRAN and can be installed using `install.packages("TDbook")`. 


Package names in this book are formatted as bold text (*e.g.*, `r Biocpkg("ggtree")`), and function names are followed by parentheses (*e.g.*, `treeio::read.beast()`). The double-colon operator (`::`) means accessing an object from a package. 

